Complementing theory on adolescent emotion regulation using  machine learning

### 2022-03-08 manuscript/manuscript.Rmd
reviewed changes manuscript
Updated manuscript, according to review of prior changes.

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to not only complete the cycle, but to enable tests confirmatory prediction following from exploratory approaches.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
Parenting behavior is another promising target for intervention,
as indeed it has historically been.
Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
By contrast, promoting positive parenting practices may have limited potential for impact or even been detrimental.
With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
another meta-analysis called these findings into question,
citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
<!-- Another study Even when negative parenting has been assessed as outcomes, -->
<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
the ultimate benefit for children,
in particular for adolescents' emotion regulation,
is less studied.
The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
that negative stimuli are more psychologically impactful than positive ones,
which has also been found in children [@vaishNotAllEmotions2008].
Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
and including children's outcomes, including emotion regulation, into consideration.
Simply put, the present results suggest that the best advice for parents might be to support their individuating adolescents' autonomy,
refrain from overly invasive or negative parenting practices,
and model problem solving behavior when conflicts arise.

@@

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world. The WEIRD nature of the sample constrict us in generalizing beyond such populations from these findings.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to enable easy replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.

>>>>

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to complete the cycle.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
Parenting behavior is another promising target for intervention,
as indeed it has historically been.
Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
By contrast, promoting positive parenting practices may have limited potential for impact.
With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
another meta-analysis called these findings into question,
citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
<!-- Another study Even when negative parenting has been assessed as outcomes, -->
<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
the ultimate benefit for children,
in particular for adolescents' emotion regulation,
is less studied.
The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
that negative stimuli are more psychologically impactful than positive ones,
which has also been found in children [@vaishNotAllEmotions2008].
Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
and including children's outcomes, including emotion regulation, into consideration.
Simply put, the present results suggest that the best advice for parents might be to support their individuating adolescents' autonomy,
refrain from overly invasive or negative parenting practices,
and model problem solving behavior when conflicts arise.

@@

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous sample from the Netherlands.
This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to enable easy replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.
